---
title: "PM10 air quality with swedishairquality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PM10 air quality with swedishairquality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

**swedishairquality** provides PM10 readings for six Swedish cities: Linköping,
Stockholm, Malmö, Gothenburg, Luleå and Umeå. PM10 is particulate matter with a
diameter of 10 µm or less (smoke, dust, soot, salts, acids and metals), formed
directly or indirectly from traffic and industrial emissions.

The package was written against SMHI's 52North REST API, which no longer
returns data at its original address. It therefore ships **`pm10_sample`, a
simulated dataset** (not real measurements) so that everything works offline.

## Getting data

`aqiGet()` returns a list of six data frames, in the order Linköping, Stockholm,
Malmö, Gothenburg, Luleå and Umeå, each with a `timestamp` and a `value`
column.

```{r}
library(swedishairquality)
pm10 <- aqiGet()
head(pm10[[1]])   # Linköping
```

```{r}
malmo <- pm10[[3]]
plot(malmo, type = "l", xlab = "Time", ylab = "PM10 (µg/m³)", main = "Malmö (simulated)")
abline(h = mean(malmo$value), col = "red")
```

To download live data, point the package at a working 52North timeseries API:

```{r eval=FALSE}
options(swedishairquality.base_url = "https://<host>/52North/api/v1")
pm10 <- aqiGet("live", months = 3)
```

## Shiny app

An interactive dashboard lets you pick a city and see its PM10 series with the
period mean:

```{r eval=FALSE}
run_app()
```
